\section{Conclusion}~\label{sec:conclusion}

The aim of this project was to analyse the performance of \acrshort{gp} based
controllers for use in longer lasting implementations, where differences in
building behaviour become important compared to the initially available data.

{\color{red}
First, the performance of a classical \acrshort{gp} model trained on 5 days
worth of experimental data was analysed. This model turned out to be unable to
correctly extrapolate building behaviour as the weather changed throughout the
year.
}

First, the performance of a classical \acrshort{gp} model trained on five days
worth of experimental data was analysed. Initially, this model performed very
well both in one step ahead prediction and multi-step ahead simulation over new,
unseen, data. With the change in weather, however, the model shifted  

Several \acrshort{svgp} implementations were then analysed. They turned out to
provide important benefits over the classical models, such as the ability to
easily scale when new data is being added and the much reduced computational
effort required. They do however present some downsides, namely increasing the
number of hyperparameters by having to choose the number of inducing locations,
as well as performing worse than then classical \acrshort{gp} implementation
given the same amount of data.

%Finally, the possible improvements to the current implementations have been
%addressed, noting that classical \acrshort{gp} implementations could also be
%adapted to the \textit{learning control} paradigm, even if their implementation
%could turn out to be much more involved and more computationally expensive than
%the \acrshort{svgp} alternative.


\subsection{Further Research}~\label{sec:further_research}

Section~\ref{sec:results} has presented and compared the results of a full-year
simulation for a classical \acrshort{gp} model, as well as a few incarnations of
\acrshort{svgp} models. The results show that the \acrshort{svgp} have much
better performance, mainly due to the possibility of updating the model
throughout the year. The \acrshort{svgp} models also present a computational
cost advantage both in training and in evaluation, due to several approximations
shown in Section~\ref{sec:gaussian_processes}.

Focusing on the \acrlong{gp} models, there could be several ways of improving
its performance, as noted previously: a more varied identification dataset and
smart update of a fixed-size data dictionary according to information gain,
could mitigate the present problems.

Using a Sparse \acrshort{gp} without replacing the maximum log likelihood
with the \acrshort{elbo} could improve performance of the \acrshort{gp} model at
the expense of training time.

An additional change that could be made is inclusion of the most amount of prior
information possible through setting a more refined kernel, as well as adding
prior information on all the model hyperparameters when available. This approach
however goes against the `spirit' of black-box approaches, since significant
insight into the physics of the plant is required in order to properly model and
implement this information.

On the \acrshort{svgp} side, several changes could also be proposed, which were
not properly addressed in this work. First, the size of the inducing dataset was
chosen experimentally until it was found to accurately reproduce the manually
collected experimental data. In order to better use the available computational
resources, this value could be found programmatically in a way to minimize
evaluation time, while still providing good performance. Another possibility is
the periodic re-evaluation of this value when new data comes in, since as more
and more data is collected the model becomes more complex, and in general more
inducing locations could be necessary to properly reproduce the training data.

Finally, none of the presented controllers take into account occupancy rates or
adapt to possible changes in the real building, such as adding or removing
furniture, deteriorating insulation and so on. The presented update methods only
deals with adding information on behaviour in different state space regions, i.e
\textit{learning}. Additionally, their ability to \textit{adapt} to changes in
the actual plant's behaviour should be further addressed.

\section*{Acknowledgements}

I would like to thank Manuel Koch for the great help provided during the course
of the project starting from the basics on CARNOT modelling, to helping me
better compare the performance of different controllers, as well as Prof.\ Colin
Jones, whose insights were always very guiding, while still allowing me to
discover everything on my own.

\clearpage
